---
title: "ppl_hhd"
output: html_document
date: "2024-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(yaml)
library(data.table)
library(dplyr)
library(datasets) # for state names
library(vroom)
```


```{r}
dict <- yaml.load_file("R:\\CR4230\\1940_2.0\\us1940b_usa.yml")
dict <- bind_rows(dict$variables)
dict_race <- read.csv("dict_race.csv")
dict_hispan <- read.csv("dict_his.csv")
dict_bpl <- read.csv("dict_bpl.csv")
dict_city <- read.csv("dict_city.csv")
dict_statefip <- read.csv("dict_state.csv")
```

Simplified Function for Column Matching
```{r}
process_ppl <- function(df)
{
  cols_to_keep <- c("SERIALP", "RACE", "BPL", "HISPAN", "NAMEFRST", "NAMELAST")
  cols_to_match <- c("RACE", "BPL", "HISPAN")
  
  df <- df[, (names(df) %in% cols_to_keep)]
  for (col in cols_to_match)
  {
    # dict_temp <- bind_rows(dict$values)[dict$name == col, ]
    dict_temp <- get(tolower(paste0('dict_', col)))
    df[, col] <- dict_temp$label[match(df[, col], dict_temp$value)]
    
    # if (sum(is.na(df[, col])) != 0)
    # {
    #   print(paste("WARNING:", sum(is.na(df[, col])), 
    #               "records in", col, "don't have a match!"))
    # }
  }
  
  df$RACE <- ifelse(df$HISPAN=='Hispanic', 'Hispanic', 
                    paste('Non-Hispanic', df$RACE))
  df$HISPAN <- NULL # remove HISPAN column
  df$NAMEFRST <- sub("^(\\w+)\\s.*", "\\1", df$NAMEFRST)
  return(df)
}
```


```{r}
process_hhd <- function(df)
{
  cols_to_keep <- c("SERIAL", "STATEFIP", "WARD", "CITY")
  cols_to_match <- c("STATEFIP", "CITY")
  
  df <- df[, (names(df) %in% cols_to_keep)]
  for (col in cols_to_match)
  {
    # dict_temp <- bind_rows(dict$values)[dict$name == col, ]
    dict_temp <- get(tolower(paste0('dict_', col)))
    df[, col] <- dict_temp$label[match(df[, col], dict_temp$value)]
    
    # if (sum(is.na(df[, col])) != 0)
    # {
    #   print(paste("WARNING:", sum(is.na(df[, col])), 
    #               "records in", col, "don't have a match!"))
    # }
  }
  
  return(df)
}
```


```{r}
process_both <- function(df_ppl, df_hhd)
{
  df <- left_join(df_ppl, df_hhd, by = c("SERIALP" = "SERIAL"))
  df$SERIALP <- NULL
  return(df)
}
```


Read, Process, and Join
```{r, message=F, warning=F}
processed_list <- list()
path <- file.path('S:', 'CR4230', 'Projects', 
                  'Project Collaboration wx233_CR4230')

# potentially using matrix or data.table to increase efficiency
# the partial datasets should also be in parquet/lighter format
start <- proc.time()
for (i in 1:109)
{
  ppl <- fread(paste0(path, '/census_people_', i,'.csv'))
  ppl <- process_ppl(data.frame(ppl))
  hhd <- fread(paste0(path, '/census_households_', i,'.csv'))
  hhd <- process_hhd(data.frame(hhd))
  processed_list[[i]] <- process_both(ppl, hhd) 
  print(paste(i, 'files completed...'))
}
print(proc.time() - start)
```
Files with mismatches:
PPL_28: [1] "WARNING: 6 records in RACE don't have a match!"
        [1] "WARNING: 534 records in BPL don't have a match!"
PPL_47: [1] "WARNING: 4 records in BPL don't have a match!"
PPL_62: [1] "WARNING: 1 records in BPL don't have a match!"
PPL_68: [1] "WARNING: 1 records in BPL don't have a match!"
PPL_82: [1] "WARNING: 1 records in BPL don't have a match!"
PPL_86: [1] "WARNING: 1 records in BPL don't have a match!"
PPL_89: [1] "WARNING: 2 records in BPL don't have a match!"
PPL_98: [1] "WARNING: 3 records in BPL don't have a match!"


Combine and Write Locally
```{r}
start <- proc.time()
output <- rbindlist(processed_list)
print(proc.time() - start)
```
```{r}
# fwrite(setDT(output), "1940_census_cleaned.parquet", row.names = F)
# write.csv(output, "1940_census_cleaned.csv", row.names = F)
# saveRDS(output, file = "1940_census_cleaned.rds")
fwrite(setDT(output), file.path(path, '1940_census_cleaned.parquet'), row.names = F)
```
```{r, message=F}
# start <- proc.time()
# output <- vroom("1940_census_cleaned.parquet")
# print(proc.time() - start)
```