---
title: "scratch"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(yaml)
library(data.table)
library(dplyr)
# library(writexl)
library(openxlsx)
```


```{r}
dict <- yaml.load_file("R:\\CR4230\\1940_2.0\\us1940b_usa.yml")
dict <- bind_rows(dict$variables)
```

```{r}
start <- proc.time()
ppl39 <- read.csv("S:\\CR4230\\Projects\\Project Collaboration wx233_CR4230\\census_people_39.csv")
proc.time() - start
# ppl40 <- read.csv("S:\\CR4230\\Projects\\Project Collaboration wx233_CR4230\\census_people_40.csv")
# ppl41 <- read.csv("S:\\CR4230\\Projects\\Project Collaboration wx233_CR4230\\census_people_41.csv")
start <- proc.time()
hhd39 <- read.csv("S:\\CR4230\\Projects\\Project Collaboration wx233_CR4230\\census_households_39.csv")
proc.time() - start
```


Match Race (hard code version)
```{r}
# race_code <- c(seq(100, 150, 10), 200, 210, 300:399, 400, 410, 420, 500, 600, 700)
# race_list <- c("White", "Spanish White", "Blank (White)", "Portuguese", "Mexican", 
#                "Puerto Rican", "Black/Negro", "Mulatto", 
#                rep("American Indian/Alaskan Native", 100),
#                "Chinese", "Taiwanese", "Chinese and Taiwanese", "Japanese", "Filipino",
#                "Other Race")

# ppl39$RACE <- race_list[match(ppl39$RACE, race_code)]
```

Match Race (yml version)
```{r}
dict_race <- bind_rows(dict$values)[dict$name == "RACE",]
ppl39$RACE <- dict_race$label[match(ppl39$RACE, dict_race$value)]
table(ppl39$RACE)
if (sum(is.na(ppl39$RACE)) == 0) {print("All Matched!")}

ppl39 <- read.csv("S:\\CR4230\\Projects\\Project Collaboration wx233_CR4230\\census_people_28.csv")
missing_idx <- which(is.na(ppl39$BPL))
```

Match Marital Status (yml version)
```{r}
dict_marst <- bind_rows(dict$values)[dict$name == "MARST",]
ppl39$MARST <- dict_marst$label[match(ppl39$MARST, dict_marst$value)]
if (sum(is.na(ppl39$MARST)) == 0) {print("All Matched!")}
```
Match Birthplace (yml version)
```{r}
dict_bpl <- bind_rows(dict$values)[dict$name == "BPL",]
ppl39$BPL <- dict_bpl$label[match(ppl39$BPL, dict_bpl$value)]
if (sum(is.na(ppl39$BPL)) == 0) {print("All Matched!")}

# p <- ppl39[,c("BPL", "BPLSTR")]
# View(p[p$BPL!=p$BPLSTR,]) # mismatches between BPL and BPLSTR
```

Match Nativity (yml version)
```{r}
dict_nat <- bind_rows(dict$values)[dict$name == "NATIVITY",]
ppl39$NATIVITY <- dict_nat$label[match(ppl39$NATIVITY, dict_nat$value)]
if (sum(is.na(ppl39$NATIVITY)) == 0) {print("All Matched!")}
```

Match Citizen Status (yml version)
```{r}
dict_cit <- bind_rows(dict$values)[dict$name == "CITIZEN",]
ppl39$CITIZEN <- dict_cit$label[match(ppl39$CITIZEN, dict_cit$value)]
if (sum(is.na(ppl39$CITIZEN)) == 0) {print("All Matched!")}
```

Match Hispanic Race (yml version)
```{r}
dict_his <- bind_rows(dict$values)[dict$name == "HISPAN",]
ppl39$HISPAN <- dict_his$label[match(ppl39$HISPAN, dict_his$value)]
if (sum(is.na(ppl39$HISPAN)) == 0) {print("All Matched!")}
```

Match Mother Tongue (yml version)
```{r}
dict_mt <- bind_rows(dict$values)[dict$name == "MTONGUE",]
ppl39$MTONGUE <- dict_mt$label[match(ppl39$MTONGUE, dict_mt$value)]
if (sum(is.na(ppl39$MTONGUE)) == 0) {print("All Matched!")}
```


Simplified Function for Column Matching
```{r}
process_df <- function(df, cols_to_match, cols_to_drop)
{
  df <- df[, !(names(df) %in% cols_to_drop)]
  
  for (col in cols_to_match)
  {
    dict_temp <- bind_rows(dict$values)[dict$name == col, ]
    df[, col] <- dict_temp$label[match(df[, col], dict_temp$value)]
    
    if (sum(is.na(df[, col])) != 0)
    {
      print(paste("WARNING:",
                  sum(is.na(df[, col])), "records in", col, "don't have a match!"))
    }
  }
  
  # Next steps: collapse the hispanic col, identify col to drop, the join tables
  
  return(df)
}


ppl39 <- read.csv("S:\\CR4230\\Projects\\Project Collaboration wx233_CR4230\\census_people_39.csv")
cols_to_match <- c("RACE", "MARST", "BPL", "NATIVITY", "CITIZEN", 
                   "HISPAN", "MTONGUE")
cols_to_drop <- c()
ppl39 <- process_df(ppl39, cols_to_match, cols_to_drop)
```


For Kate: Column Categories
```{r}
wb <- createWorkbook()

addWorksheet(wb, sheetName = "RACE")
writeData(wb, sheet = 1, x = dict_race)

addWorksheet(wb, sheetName = "HISPAN")
writeData(wb, sheet = 2, x = dict_his)

race_hispan <- expand.grid(RACE = dict_race$label, HISPAN = dict_his$label)
count <- table(match(paste(ppl39$RACE, ppl39$HISPAN),
                      paste(race_hispan$RACE, race_hispan$HISPAN)))
idx <- as.numeric(names(result))
race_hispan$Count <- 0
race_hispan$Count[idx] <- count
race_hispan <- race_hispan[order(race_hispan$Count, decreasing = T), ]
addWorksheet(wb, sheetName = "RACE-HISAN")
writeData(wb, sheet = 3, x = race_hispan)

addWorksheet(wb, sheetName = "BPL")
writeData(wb, sheet = 4, x = dict_bpl)

addWorksheet(wb, sheetName = "CITIZEN")
writeData(wb, sheet = 5, x = dict_cit)

# saveWorkbook(wb, "Data Dictionary for Kate.xlsx")
```


Export Dict for RACE for Manual Recoding
Will recode HISPAN/BPL inline
```{r}
# dict_ward <- bind_rows(dict$values)[dict$name == "WARD",]
# write.csv(dict_ward, "dict_ward.csv", row.names = F)
# dict_state <- bind_rows(dict$values)[dict$name == "STATEFIP",]
# write.csv(dict_state, "dict_state.csv", row.names = F)
# dict_city <- bind_rows(dict$values)[dict$name == "CITY",]
# write.csv(dict_city, "dict_city.csv", row.names = F)

# write.csv(dict_his, "dict_his.csv", row.names = F)
# write.csv(dict_his, "dict_his.csv", row.names = F)
# write.csv(dict_his, "dict_his.csv", row.names = F)
```



